6.878 Final Report

### 2013-12-10 report/paper.tex
final submission

  Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering). We make a brief mention of visualization and annotation for completion.

@@

The RNASeq dataset posted on GEO is the output of topHat (~120GB) and we performed the following preprocessing.
\begin{description}
 \item a. Cufflinks analysis to derive gene expression abundance in FPKMs
 \item b. Filtering,Quantile normalization,vector normalization of the cufflinks output.
 \item c. We then created multiple views of this expression data to get a feel for how the
		data looks.
\end{description}

@@

\begin{table}
\caption{Some sample scoring matrices alignment: least strict (allowing matches between drops and rises in signal for potention downregulation effects), more strict (giving mismatches between rises and drops), most strict (penalizing mismatches). The gap penalty depends on the time offset between two time points, increasing with time differences.}
\label{tab:scoring}
\begin{center}
\begin{tabular}{c | ccc}
  & R & D & S\\ \hline

@@

Given the distance matrix generated by alignment, we were able to cluster similar sequences.
Our clustering was based on the alignment scores generated above as distance.
One question that needed to be resolved was how determine the optimal number of
clusters.We have tried the following two approaches towards this
a. Pairwise clustering until we reach a threshold of the alignment score between
two clusters.
b. Affinity propagation based clustering to automatically determine the number of
clusters.The pairwise heirarchical clustering with an emperical value of threshold based
out of experiments gives us fewer clusters than the affinity propogation . However the clusters 
generated by affinity propogation appear to be more cohesive

@@

@@

While, visually, the clusters showed alignment for GSE33029 dataset , the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure. However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means (Fig. \ref({fig:enrichment}). It shows that, to some degree, our alignment and clustering technique performs better than without alignment.\\

@@

We then used \verb!DAVID! Bioinformatics Database web service to find functional groups within each cluster, as a means of describing each cluster, with $kappa=70$ and $linkage = 0.50$. We were able to generate functional groups for several clusters (Fig. \ref({fig:enrichment}).
The \verb!DAVID! clustering provided functional groups within each cluster, whereas corrected Fischer's Exact test tries to summarize an entire cluster. Since we were able to find significant groups within each cluster, but not for an entire cluster, this may indicate that our clustering is too general or that the clusters are too large.

@@

We have been unable to pinpoint the cause for the lack of enrichment annotations for our clusters on the GSE33029 dataset.We suspected that we are missing a noise correction step for the GSE33029 dataset that may be resulting in no enrichment showing up and reanalyzed with additional normalizations in no vain. We suspect the authors may have not fully specified the steps they took to preprocessing this dataset and we may be missing some correction step.

To check if our approach works we decided to rerun our analysis on two additional and independent timecourse datasets from GEO (GSE19392 GSE675) . With clusters generated from these datasets we do significant enrichment numbers showing up for pathways and other annotation terms. The KEGG pathway results for a few of our clusters are as shown below.

\begin{figure}[H]
\centering
\includegraphics[scale=0.65]{3Clusters_PathwayCombined.png}
\caption{KEGG Pathway analysis on 3 clusters from GSE19392 dataset using david.}
\label{fig:KEGGPathway.}
\end{figure}
\subsection{Visualization}

In order to make expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3! that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID! gene groups by cluster (Fig. \ref({fig:enrichment}).
\section{Future Goals}
We would re-do the pre-processing step or attempt to get the pre-processed dataset before reanalyzing our data.

@@

\section{Comparison with original proposal} 
We managed to complete everything that we proposed, including our alignment technique, several forms of clustering (Affinity Propogation, Hierarchical, K-means), several forms of enrichment analysis (DAVID's function annotations, pulling OMIM text, DAVID's gene grouping, Fischer's Exact test with multiple hypothesis corrections). We also provided a visualization tool as part of our combined class project.

@@

Dalesh: Working on the project was an enjoyable experience , a few highlights are as follows
a. Opportunity to work in a team and have a mentor to guide us.
b. Applying our learning on a couple of real datasets gave additional insights.
c. The two interim feedback checkpoints

\section{Commentary on peer-review process}
Iris: I thought that it was good to get feedback, but at the same time, our proposal was missing a lot of detail on purpose because we were not sure at the time exactly on what we were going to do, and a lot of the feedback was on lack of detail, which was not that helpful. It was nice to see other people's ideas.
Dalesh: The peer review process was fun . And i agree with Iris's comment on the feedback we received , i recollect giving ratings on the table without having completely understood what the authors were trying to communicate.

@@

Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.
Dalesh: Most of the strategy and design was arrived at by brainstorming together , my specific implementation contributions were in data preprocessing on the amazon EC2 instances and clustering , I was responsible for processing and analyzing the GSE19392 dataset , we split the analysis of the GSE33029 dataset . Iris also did the implementations for the user interface.

>>>>

  Note : While our overall project was done as part of two of our classes , in this report we highlight and detail the methods learnt and relevant to this class (Pre-processing , alignment , clustering, Enrichment analysis). We make a brief mention of visualization and annotation for completion.

@@

The RNASeq dataset ( ~120 gb ) posted on GEO is the output of topHat and we had to perform the following preprocessing.
\begin{description}
 \item a. Cufflinks analysis to derive gene expression abundance in FPKMs
 \item b. Filtering,Quantile normalization,vector normalization of the cufflinks output.
 \item c. We then created multiple views of this expression data to get a feel for how the
		data looks.
\end{description}

@@

\begin{table}[H]
\caption{Some sample scoring matrices alignment: least strict (allowing matches between drops and rises in signal for potention downregulation effects), more strict (giving mismatches between rises and drops), most strict (penalizing mismatches). The gap penalty depends on the time offset between two time points, increasing with time differences.}
\label{tab:scoring}
\begin{center}
\begin{tabular}{c | ccc}
  & R & D & S\\ \hline

@@

Given the distance matrix generated by alignment, we were able to cluster similar sequences.
Our clustering was based on the alignment scores generated above as distance.
One question that needed to be resolved was how determine the optimal number of
clusters.We have tried the following two approaches towards this
a. Pairwise clustering until we reach a threshold of the alignment score between
two clusters.
b. Affinity propagation based clustering to automatically determine the number of
clusters.The pairwise heirarchical clustering with an emperical value of threshold based
out of experiments gives us fewer clusters than the affinity propogation . However the clusters 
generated by affinity propogation appear to be more cohesive.

@@

We initially reused the pset versions of our alignment and clustering code but had to make more efficient versions to run through the larger dataset.

@@

While, visually, the clusters showed alignment for GSE33029 dataset , the p-value matrices indicated no significant enrichment in our clusters after correcting for multiple comparisions with the Benjamini-Hochberg procedure. However, in order to get some quantitative measure of performance, we took sums of the p-value matrices generated from enrichment analysis for our technique for the GSE675 dataset and normalized over the same value for clusters generated by k-means. It shows that, to some degree, our alignment and clustering technique performs better than without alignment.\\

To further check if our approach works we decided to rerun our analysis on two additional and independent timecourse datasets from GEO (GSE19392 GSE675). With clusters generated from these datasets we do see significant enrichment numbers showing up for pathways and other annotation terms. The KEGG pathway results for a few of our clusters on the above datasets are as shown below. We havent yet related these results to the context of the original studies. Additional time is needed though to accurately conclude on the Enrichment analysis to pinpoint the cause for the lack of enrichment annotations for our clusters on the GSE33029 dataset.We suspected that we are missing a noise correction step for the GSE33029 dataset that may be resulting in no enrichment showing up and reanalyzed with additional normalizations in no vain. We also suspect the authors may have not fully specified the steps they took to preprocessing this dataset and we may be missing some correction step.

\begin{figure}[H]
\centering
\includegraphics[scale=0.65]{3Clusters_PathwayCombined.png}
\caption{KEGG Pathway analysis on 3 clusters from GSE19392 dataset using david.}
\label{fig:KEGGPathway.}
\end{figure}

@@

We then used \verb!DAVID! Bioinformatics Database web service to find functional groups within each cluster, as a means of describing each cluster, with $kappa=70$ and $linkage = 0.50$. We were able to generate functional groups for several clusters.
The \verb!DAVID! clustering provided functional groups within each cluster, whereas corrected Fischer's Exact test tries to summarize an entire cluster. Since we were able to find significant groups within each cluster, but not for an entire cluster, this may indicate that our clustering is too general or that the clusters are too large.

@@

\subsection{Visualization}

In order to make the expression trend analysis easier from a clinical standpoint, we created a browser-based visualization tool using \verb!d3! that provides normalized/denormalized expression trends by cluster, dendrograms for cluster annotation, and word clouds for a text summary of \verb!DAVID! gene groups by cluster.

\section{Future Goals}
We would redo the analysis on GSE33029 dataset , to identify the root cause of clusters not showing corresponding enrichment results.

@@

\section{Comparison with original proposal} 
Our original proposal had many items we were going to exlplore and choose from.
We managed to complete everything that we proposed on our peer review response , including our alignment technique, several forms of clustering (Affinity Propogation, Hierarchical, K-means), several forms of enrichment analysis (DAVID's function annotations, pulling OMIM text, DAVID's gene grouping, Fischer's Exact test with multiple hypothesis corrections). We also provided a visualization tool as part of our combined class project.

@@

Dalesh: Working on the project was an enjoyable experience , a few highlights are as follows
a. Opportunity to work in a team and have a mentor to guide us.
b. Applying our learning on a couple of real datasets gave additional insights.
c. Structuring of the project activity with milestones starting early in the class with two interim feedback checkpoints were helpful as well.

\section{Commentary on peer-review process}
Iris: I thought that it was good to get feedback, but at the same time, our proposal was missing a lot of detail on purpose because we were not sure at the time exactly on what we were going to do, and a lot of the feedback was on lack of detail, which was not that helpful. It was nice to see other people's ideas.

Dalesh: The peer review discussion was fun but I agree with Iris's comment on the feedback we received , i recollect giving ratings on the table without having completely understood what the authors were trying to communicate and in many cases the initial proposals were in a premature stage.

@@

Iris: I did some of the alignment, enrichment analysis, and clustering and helped get jobs running on Broad clusters. I also did the analysis and processing for the GSE675 dataset.

Dalesh: Most of the strategy and design was arrived at by brainstorming together , my specific implementation contributions were in data preprocessing on the amazon EC2 instances and implementations for clustering , I was responsible for processing and analyzing the GSE19392 dataset , we split the analysis of the GSE33029 dataset . Iris also did the implementations for the user interface.
